---
title: "World Edible Plants: Bayesian Assessment of Edibility"
output: html_document
bibliography: bib/packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(brms)
library(bayesplot)
library(HDInterval)

# Load the custom edibility classification function
source("functions/check_edibility_wikipedia.R")

# Load the dataset from the data folder
species_data <- read.csv("data/randomsample200species_and_synonims.csv")
```

# Introduction

This study aims to estimate the proportion of edible vascular plant species globally. While estimates vary from fewer than 7,000 to over 30,000 edible species, these numbers depend on definitions influenced by culture, toxicity, and available processing technologies.

We follow a method proposed by [Colleague's Name], focusing on a randomly selected sample of 200 plant species to assess edibility.

# Methods

## Defining Edibility Criteria

We categorize each species based on:

- **Evidence Strength**:
  - Number of independent ethnobotanical reports.
  - Lack of toxicity evidence.
- **Processing Requirements**:
  - Edible raw (e.g., *Brassica oleracea*).
  - Edible after cooking (e.g., apples).
  - Edible after further processing (e.g., *Solanaceae* spp.).
- **Quantity Consideration**:
  - Differentiate between food additives and staples (e.g., spices like nutmeg).

## Sample and Data Collection

The dataset contains 200 species, including 619 synonyms. Assuming a 10% population proportion of edible species, this sample size offers a 95% confidence interval with a ±4.16% margin of error.

### Literature Search Strategy

- **Search Queries**:
  - "species name" AND "food"
  - "species name" AND "edible"
- **Databases**:
  - Wikipedia (automated extraction)
  - Scopus (manual follow-up for unclassified species)
  - Consensus API (automated search for scientific literature)

```{r consensus-api-query}
library(httr)
library(jsonlite)

# Define function to query the Consensus API for edibility information
query_consensus <- function(species_name) {
  api_url <- "https://api.consensus.app/v1/search"
  api_key <- "YOUR_API_KEY"  # Replace with your actual API key
  query <- paste0("\"", species_name, "\" AND (food OR edible OR nutrition)")

  response <- tryCatch({
    httr::POST(
      api_url,
      add_headers(Authorization = paste("Bearer", api_key)),
      encode = "json",
      body = list(query = query)
    )
  }, error = function(e) return(NULL))

  if (is.null(response) || http_error(response)) {
    return("unknown")
  }

  # Parse response
  content <- content(response, as = "parsed", type = "application/json")
  
  if (length(content$results) == 0) {
    return("unknown")
  }
  
  # Extract relevant text
  text_data <- tolower(paste(unlist(content$results), collapse = " "))
  
  # Define keywords for classification
  edible_keywords <- c("edible", "consumed", "food", "eaten", "nutrition")
  
  if (any(str_detect(text_data, edible_keywords))) {
    return("edible")
  } else {
    return("not_edible")
  }
}

# Apply the function to the dataset
species_data$consensus_edibility <- sapply(species_data$scientificName, query_consensus)

# Save updated dataset
write.csv(species_data, "data/updated_species_with_consensus.csv", row.names = FALSE)
``````{r initialize-edibility-status}
# Placeholder: Assume initial unknown edibility status
species_data$edibility_status <- NA  # Will be updated after literature search

# Display initial dataset
head(species_data)
```

# Bayesian Modeling of Edibility

## Bayesian Model Specification and Posterior Inference

This analysis utilizes several R packages for modeling and visualization. Specifically, we employ `brms` for Bayesian modeling (@R-brms), `bayesplot` for diagnostic visualization (@R-bayesplot), `tidyverse` for data manipulation (@R-tidyverse), and `HDInterval` for computing highest density intervals (@R-HDInterval). These packages provide robust tools for implementing Bayesian hierarchical models and interpreting the resulting posterior distributions.

## Model Specification

We fit a Bayesian logistic regression model to estimate the proportion of edible species based on the available evidence.

```{r bayesian-model}
# Set non-informative priors
alpha_prior <- 1
beta_prior <- 1

# Placeholder values assuming future search updates
species_data$edibility <- ifelse(is.na(species_data$edibility_status), 0, 1)  # 1 for edible, 0 otherwise

total_edible <- sum(species_data$edibility)
total_species <- nrow(species_data)

# Posterior update
alpha_post <- alpha_prior + total_edible
beta_post <- beta_prior + total_species - total_edible

# Calculate credible intervals
credible_interval <- hdi(rbeta(10000, alpha_post, beta_post), credMass = 0.95)

list(
  mean_estimate = alpha_post / (alpha_post + beta_post),
  credible_interval = credible_interval
)
```

# Posterior Predictive Check

```{r posterior-predictive-check}
# Diagnostic to assess model fit
pp_check(brm(edibility ~ 1, data = species_data, family = bernoulli()))
```

# References
